---
title: "Clean_HPA"
format: html
editor: visual
---

```{r}
#install.packages("vroom")
#install.packages("installr")
library(vroom)
library(tidyverse)
library(here)
library(readxl)
library(janitor)

#984 MB of data hahaha
list_all<- list.files(
  here("data","hpa_scraped_metadata_files"),
           full.names = TRUE ) 
length(list_all)
Ophelia_data <- read.csv(here("data", "10K_Single_Cell_Filtered(in).csv"))

files_col_name <- c("gene", "tissue", "antibody","image_link","info_id", "gender", "age", "tissue_detail", "sample_category", "patient_id", "cell_type", "expression_level_category", "reliability", "quantity","location")
#worked but there are files with no content like files 361 and 363 edit: solved 

all_metadata <- vroom::vroom(
  list.files(here("data","hpa_scraped_metadata_files"),
           full.names = TRUE),
  delim = NULL,
  col_names = files_col_name,
  progress = TRUE,
  id = "source_file")
```

Barplot showing exp_level between "tissue", want to see not detected across everything

```{r}
all_metadata|> 
  ggplot(aes(x = tissue, fill = expression_level_category)) +
  geom_bar(position = "fill") +
  labs(title = "Protein Expression Levels by Tissue",
       x = "Tissue Type", y = "Proportion") +
  theme_minimal()
```

Now let's explore our patients, how old they are, what types of tissues they have, gender divide

```{r}
all_metadata |> 
  count(age, sort = TRUE)

all_metadata |> 
  filter(!is.na(age)) |> 
  distinct(patient_id, .keep_all = TRUE) |> 
  ggplot() + aes(x = age, fill = gender) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  theme_minimal() +
  labs(title = "Age Distribution by Gender")
ggsave(here("output", "AgeDistributionByGender.pdf"))
```

Cross tables, using tabyl to make how many times x variable appear in each tissue

```{r}
cross_gene_tissue <- all_metadata |> 
  tabyl(gene, tissue) |> 
  adorn_totals("both") 
cross_gene_tissue_prop <- all_metadata |> 
  tabyl(gene, tissue) |> 
  adorn_percentages("row") |> 
  adorn_pct_formatting()
cross_anti_reliability <- all_metadata |> 
  tabyl(antibody, reliability) |> 
  adorn_totals("both")
cross_anti_reliability_prop <- all_metadata |> 
  tabyl(antibody, reliability) |> 
  adorn_percentages("row") |> 
  adorn_pct_formatting()


#an example of three way cross 
cross3way_rel_patient_sampleCat <- all_metadata |> 
  group_by(reliability) |> 
  summarise(
    table = list(tabyl(cur_data(), patient_id, sample_category)))
```

Ophelia's data, general exploration

```{r}

Ophelia_data |> 
  count(Patient_ID, sort = TRUE)
#less patients only 954 instead of 1,441
all_metadata |> 
  count(patient_id, sort = TRUE)
#let's take at look at a patient of interest 
Ophelia_data |> 
  filter(Patient_ID == "2523")
#130 data points for 2523 is it all the same? let's see also important note for her staining is exp_lvl_cat and reliability is intensity which makes sense 
all_metadata |> 
  filter(patient_id == "2523") |> 
  filter(tissue_detail == "Heart (T-32000)") |> 
  ggplot() + aes(x = reliability) +
  geom_bar() +
  theme_minimal()
Ophelia_data |> 
  filter(Patient_ID == "2523") |> 
  filter(Tissue_Location == "Heart (T-32000)") |> 
  ggplot() + aes(x = Intensity) +
  geom_bar() +
  scale_fill_brewer() +
  theme_minimal()
```

To answer the question, Is patient 2523's normal tissue as reliable as any other patient? -\> Does the distribution of reliability scores in normal tissue differ between patients?

```{r}
all_metadata |> 
  filter(patient_id %in% c("2523","3731")) |> 
  select(sample_category, reliability, patient_id) |> 
  count(reliability)
#both are only normal tissue, we still see varible reliability
  
class(all_metadata$patient_id) 

all_metadata |> 
  filter(sample_category =="Normal tissue, NOS (M-00100)") |> 
  count(patient_id)

normal_props <- all_metadata |> 
  filter(sample_category =="Normal tissue, NOS (M-00100)") |> 
  count(patient_id,reliability) |> 
  group_by(patient_id) |> 
  mutate(prop = n / sum(n)) |> 
  ungroup()
#now i will plot per my two highest count patients 

normal_props |> 
   filter(patient_id %in% c("2523","3731")) |> 
  ggplot() + aes(x= factor(patient_id), y = prop, fill = reliability) + geom_col()

```

this is a test

```{r}
all_metadata |>
  group_by(patient_id) |>
  summarise(n_total = n(),
    has_not_detected = any(expression_level_category == "Not detected", na.rm = TRUE),
    has_other = any(expression_level_category != "Not detected", na.rm = TRUE),
    .groups = "drop") |>
  mutate(category = case_when(
      has_not_detected & has_other ~ "mixed (not detected + other result)",
      has_not_detected & !has_other ~ "only not detected",
      !has_not_detected & has_other ~ "no not detected (only real results)",
      TRUE ~ "other"))

#same but now i want to include the gene
all_metadata |>
  group_by(patient_id, gene) |>
  summarise(n_tests = n(),
    has_not_detected = any(expression_level_category == "Not detected",
                           na.rm = TRUE),
    has_other = any(expression_level_category != "Not detected",
                    na.rm = TRUE),
    n_not_detected = sum(expression_level_category == "Not detected", 
                         na.rm = TRUE),
    n_other = sum(expression_level_category != "Not detected", 
                  na.rm = TRUE),
    .groups = "drop") |>
  mutate(category = case_when(
      has_not_detected & !has_other ~ "only not detected (no retest)",
      has_not_detected & has_other ~ "not detected and a retest",
      !has_not_detected ~ "never not detected"))
```

Full Join of both datasets (missing clarification on tissues: T-2300, T-X1600 and T-X2570

```{r}
#Processed 
mergeable_metadata <- all_metadata |> 
  mutate(tissue = str_to_title(tissue),
         tissue = str_remove(tissue, "\\+.*"),
         antibody = str_remove(antibody, "Antibody ")) |> 
  rename(Protein = gene,
         Patient_ID = patient_id, 
         Tissue = tissue)

great_join <- Ophelia_data |> 
  left_join(mergeable_metadata |> 
  select(antibody, 
         Protein, 
         Patient_ID, 
         Tissue, 
         reliability, 
         sample_category), 
  by =  c("antibody", "Protein", "Patient_ID", "Tissue"))

setdiff(Ophelia_data$Tissue, mergeable_metadata$Tissue)

all_metadata |>  
  filter(str_detect(tissue_detail, "(T-23000)|(T-X1600)| (T-X2570)"))


```
