---
title: "Clean_HPA"
format: html
editor: visual
---

Sample Variability Analysis

Act 1

Libraries used:

```{r}
library(vroom)
library(tidyverse)
library(here)
library(readxl)
library(janitor)
library(irr)
library(patchwork)
```

Data loading

```{r}
files_col_name <- c("gene", "tissue", "antibody","image_link","info_id", "gender", "age", "tissue_detail", "sample_category", "patient_id", "cell_type", "expression_level_category", "reliability", "quantity","location")

all_metadata <- vroom::vroom(
  list.files(here("data","hpa_scraped_metadata_files"),
           full.names = TRUE),
  delim = NULL,
  col_names = files_col_name,
  progress = TRUE,
  id = "source_file")

HUEQ_data <- read.csv(here("data", "10K_Single_Cell_Filtered(in).csv"))

#reliability column in all_metadata is the "intensity" variable from HPA webiste
all_metata <- all_metata |> 
  rename(intensity = )


```

Barplot showing exp_level between "tissue"

```{r}
all_metadata|> 
  ggplot(aes(x = tissue, fill = expression_level_category)) +
  geom_bar(position = "fill") +
  labs(title = "Protein Expression Levels by Tissue",
       x = "Tissue Type", y = "Proportion") +
  theme_minimal()
```

Patient counts

```{r}
all_metadata |> 
  count(age, sort = TRUE)
all_metadata |> 
  count(patient_id, sort = TRUE)

all_metadata |> 
  filter(!is.na(age)) |> 
  distinct(patient_id, .keep_all = TRUE) |> 
  ggplot() + aes(x = age, fill = gender) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  theme_minimal() +
  labs(title = "Age Distribution by Gender")
ggsave(here("output", "AgeDistributionByGender.pdf"))
```

Main figure for counts

```{r}



```

Act 2

Cross tables, using tabyl to make how many times x variable appear in each tissue. Used for descriptive structre

```{r}
#Gene x tissue frequency and proportions 
cross_gene_tissue <- all_metadata |> 
  tabyl(gene, tissue) |> 
  adorn_totals("both") 

cross_gene_tissue_prop <- all_metadata |> 
  tabyl(gene, tissue) |> 
  adorn_percentages("row") |> 
  adorn_pct_formatting()

#Antibody x reliability frequency and proportions 
cross_anti_reliability <- all_metadata |> 
  tabyl(antibody, reliability) |> 
  adorn_totals("both")
cross_anti_reliability_prop <- all_metadata |> 
  tabyl(antibody, reliability) |> 
  adorn_percentages("row") |> 
  adorn_pct_formatting()


#an example of three way cros: reliability x patient x sample category 
cross3way_rel_patient_sampleCat <- all_metadata |> 
  group_by(reliability) |> 
  summarise(
    table = list(tabyl(cur_data(), patient_id, sample_category)))
```

"Not detected" category profiling per patient and gene: Classifies each patient by whether their entries are entirely "Not detected", mixed or clean.

```{r}
#Purely per patient
all_metadata |>
  group_by(patient_id) |>
  summarise(n_total = n(),
    has_not_detected = any(expression_level_category == "Not detected", na.rm = TRUE),
    has_other = any(expression_level_category != "Not detected", na.rm = TRUE),
    .groups = "drop") |>
  mutate(category = case_when(
      has_not_detected & has_other ~ "mixed (not detected + other result)",
      has_not_detected & !has_other ~ "only not detected",
      !has_not_detected & has_other ~ "no not detected (only real results)",
      TRUE ~ "other"))

#Per Patient x gene 
all_metadata |>
  group_by(patient_id, gene) |>
  summarise(n_tests = n(),
    has_not_detected = any(expression_level_category == "Not detected",
                           na.rm = TRUE),
    has_other = any(expression_level_category != "Not detected",
                    na.rm = TRUE),
    n_not_detected = sum(expression_level_category == "Not detected", 
                         na.rm = TRUE),
    n_other = sum(expression_level_category != "Not detected", 
                  na.rm = TRUE),
    .groups = "drop") |>
  mutate(category = case_when(
      has_not_detected & !has_other ~ "only not detected (no retest)",
      has_not_detected & has_other ~ "not detected and a retest",
      !has_not_detected ~ "never not detected"))
```

Do reliability distributions differ between patients, within the same tissue?

```{r}
normal_props <- all_metadata |> 
  filter(sample_category =="Normal tissue, NOS (M-00100)") |> 
  count(patient_id,reliability) |> 
  group_by(patient_id) |> 
  mutate(prop = n / sum(n)) |> 
  ungroup()
#Plot per two highest count patients 

normal_props |> 
   filter(patient_id %in% c("2523","3731")) |> 
  ggplot() + aes(x= factor(patient_id), y = prop, fill = reliability) + geom_col()

```

Data Merge & Exploration

```{r}
#Transforming data into proper formatting
mergeable_metadata <- all_metadata |> 
  mutate(tissue = str_to_title(tissue),
         tissue = str_remove(tissue, "\\+.*"),
         antibody = str_remove(antibody, "Antibody ")) |> 
  rename(Protein = gene,
         Patient_ID = patient_id, 
         Tissue = tissue)
#Joining and making proper ordinal values for relevant labels 
great_join <- HUEQ_data |> 
  left_join(mergable_metadata |> 
  select(antibody, 
         Protein, 
         Patient_ID, 
         Tissue, 
         reliability, 
         sample_category), 
  by =  c("antibody", "Protein", "Patient_ID", "Tissue")) |> 
  mutate(quantity_ord = factor(Quantity, 
                               levels = c("None", "<25%", "75%-25%", ">75%"), 
                               ordered = TRUE), 
         staining_ord = factor(Staining, 
                               levels = c("Not detected", "Low", "Medium", "High"),
                               ordered = TRUE), 
         intensity_ord = factor(Intensity, 
                                levels = c("Negative", "Weak", "Moderate", "Strong"),
                                ordered = TRUE), 
         reliability_ord = factor(reliability, 
                                  levels = c("Negative", "Weak", "Moderate", "Strong"),
                                  ordered = TRUE))

#EDIT 
unmatched_data <- setdiff(HUEQ_data$Tissue, mergable_metadata$Tissue)
unmatched_data


```

ACT3: Concordance Analysis Labels "Stain" x "Quantity

```{r}
#EDIT: EXPLAIN
concordance_StainQuant <- great_join |>
  filter(!is.na(reliability))|> 
  tabyl(staining_ord, quantity_ord) |> 
  adorn_totals("both") |> 
  adorn_percentages("row") |> 
  adorn_pct_formatting()
concordance_StainQuant

#Concordance table visualization
great_join |> 
  filter(!is.na(reliability)) |> 
  count(staining_ord, quantity_ord) |> 
  group_by(staining_ord) |> 
  mutate(prop = n / sum(n)) |> 
  ungroup() |> 
  ggplot() + aes(x = staining_ord, y = quantity_ord, fill = prop) +
  geom_tile() + geom_text(aes(label = scales::percent(prop, accuracy = 1)),
                          colour = "white", frontface = "bold") +
  labs(title = "HPA Expression Label vs Image-derived Staining %",
       y = "Image-derived Quantity (%)", 
       x = "HPA Expression Level Category", 
       fill = "Row proportion") 
ggsave(here("output", "Act3_concor_Heatmap.pdf"))



```

Kappa Value for Staining Label (check if able to use)

```{r}
kappa_stain <- great_join |> 
  filter(!is.na(reliability)) |> 
  transmute(
    rater1 = as.numeric(staining_ord),
    rater2 = as.numeric(quantity_ord)) |> 
  kappa2(weight = "squared")
kappa_stain

```

Are the reliability labels meaningful? - This is a direct label from all_metadata, basically analysis in three ways - HUEQ's data against all_metadata - HPA site labels against Image-Derived % - Is reliability as good as staining (ei. exp_lvl_cat)?

```{r}
#Do rows in reliability (all_metadata) "High" have higher concordance with HUEQ's data than rows where HPA says "Negative"?
great_join |> 
  mutate(concordant = case_when(is.na(reliability) ~ NA, 
         is.na(quantity_ord) | is.na(staining_ord) ~ NA,
         quantity_ord == "None" & staining_ord == "Not detected" ~ TRUE,
         quantity_ord == "<25%" & staining_ord == "Low" ~ TRUE,
         quantity_ord == "75%-25%" & staining_ord == "Medium" ~ TRUE,
         quantity_ord == ">75%" & staining_ord == "High" ~ TRUE,
         TRUE  ~ FALSE )) |> 
  filter(!is.na(concordant)) |> 
  group_by(reliability) |> 
  summarise(n = n(),
            n_concordant = sum(concordant),
            pct_concordant = round(n_concordant / n *100, 1),
            .groups = "drop") |> 
  arrange(desc(pct_concordant))

#Concordance table visualization of reliability 
great_join |> 
  filter(!is.na(reliability)) |> 
  count(reliability_ord, quantity_ord) |> 
  group_by(reliability_ord) |> 
  mutate(prop = n / sum(n)) |> 
  ungroup() |> 
  ggplot() + aes(x = reliability_ord, y = quantity_ord, fill = prop) +
  geom_tile() + geom_text(aes(label = scales::percent(prop, accuracy = 1)),
                          colour = "white", frontface = "bold") +
  labs(title = "HPA Reliability Label vs Image-derived Staining %",
       y = "Image-derived Quantity (%)", 
       x = "HPA Reliability", 
       fill = "Row proportion") 
ggsave(here("output", "Act3_concorReli_Heatmap.pdf"))

```

Kappa Value for Staining Label (check if able to use)

```{r}
kappa_reliability <- great_join |> 
  filter(!is.na(reliability)) |> 
  transmute(
    rater1 = as.numeric(reliability_ord),
    rater2 = as.numeric(quantity_ord)) |> 
  kappa2(weight = "squared")
kappa_reliability

```

Identifying nonconcordant observations

```{r}
nonconcordant <- great_join |> 
  mutate(concordant = case_when(is.na(reliability) ~ NA, 
         is.na(quantity_ord) | is.na(staining_ord) ~ NA,
         quantity_ord == "None" & staining_ord == "Not detected" ~ TRUE,
         quantity_ord == "<25%" & staining_ord == "Low" ~ TRUE,
         quantity_ord == "75%-25%" & staining_ord == "Medium" ~ TRUE,
         quantity_ord == ">75%" & staining_ord == "High" ~ TRUE,
         TRUE  ~ FALSE )) |>
  filter(!is.na(reliability), concordant == FALSE)

#Top 15 tissues
nonconcordant |> 
  count(Tissue, sort = TRUE) |> 
  mutate(prop_nonconcordant = n / sum(n)) |> 
  head(15) |> 
  ggplot() + aes(x = reorder(Tissue, n), y = n, fill = n) + 
  geom_col() + coord_flip() +
  scale_fill_distiller(palette = "Spectral") +
  labs(title = "Top 15 Tissues by Nonconcordance Count",
       x = "Tissue",
       y = "Number of nonconcordant rows", 
       fill = " ")
  
ggsave(here("output", "Act3_top15_noncor.pdf"))

#Checking for rate within data, sister plot 
great_join |>
  mutate(concordant = case_when(is.na(reliability) ~ NA, 
         is.na(quantity_ord) | is.na(staining_ord) ~ NA,
         quantity_ord == "None" & staining_ord == "Not detected" ~ TRUE,
         quantity_ord == "<25%" & staining_ord == "Low" ~ TRUE,
         quantity_ord == "75%-25%" & staining_ord == "Medium" ~ TRUE,
         quantity_ord == ">75%" & staining_ord == "High" ~ TRUE,
         TRUE  ~ FALSE )) |> 
  filter(!is.na(reliability), !is.na(concordant)) |>
  group_by(Tissue) |>
  summarise(n_total        = n(),
    n_nonconcordant = sum(!concordant, na.rm = TRUE),
    pct_nonconcordant = round(n_nonconcordant / n_total * 100, 1),
    .groups = "drop") |>
  filter(n_total >= 50) |>       
  slice_max(pct_nonconcordant, n = 15) |>
  ggplot(aes(x = reorder(Tissue, pct_nonconcordant), 
             y = pct_nonconcordant,
             fill = pct_nonconcordant)) +
  geom_col() +
  coord_flip() +
  scale_fill_distiller(palette = "RdYlGn", 
                       direction = -1,      
                       name = "% nonconcordant") +
  geom_text(aes(label = paste0(pct_nonconcordant, "%")),
            hjust = -0.2, size = 3.5) +
  labs(
    title    = "Top 15 Tissues by Nonconcordance Rate",
    subtitle = "Only tissues with ≥ 50 rows included",
    x        = "Tissue",
    y        = "% nonconcordant rows") +
  theme_minimal(base_size = 12) +
  expand_limits(y = 100)
ggsave(here("output", "Act3_top15_noncor_rate.pdf"))

```

Nonconcordance by antibody

```{r}

great_join |>
  mutate(concordant = case_when(is.na(reliability) ~ NA, 
         is.na(quantity_ord) | is.na(staining_ord) ~ NA,
         quantity_ord == "None" & staining_ord == "Not detected" ~ TRUE,
         quantity_ord == "<25%" & staining_ord == "Low" ~ TRUE,
         quantity_ord == "75%-25%" & staining_ord == "Medium" ~ TRUE,
         quantity_ord == ">75%" & staining_ord == "High" ~ TRUE,
         TRUE  ~ FALSE )) |> 
  filter(!is.na(reliability), !is.na(concordant)) |>
  group_by(antibody) |>
  summarise(n_total        = n(),
    n_nonconcordant = sum(!concordant, na.rm = TRUE),
    pct_nonconcordant = round(n_nonconcordant / n_total * 100, 1),
    .groups = "drop") |>
  filter(n_total >= 50) |>       
  slice_max(pct_nonconcordant, n = 15) |>
  ggplot(aes(x = reorder(antibody, pct_nonconcordant), 
             y = pct_nonconcordant,
             fill = pct_nonconcordant)) +
  geom_col() +
  coord_flip() +
  scale_fill_distiller(palette = "RdYlGn", 
                       direction = -1,      
                       name = "% nonconcordant") +
  geom_text(aes(label = paste0(pct_nonconcordant, "%")),
            hjust = -0.2, size = 3.5) +
  labs(
    title    = "Top 15 Antibodies by Nonconcordance Rate",
    subtitle = "Only Antibodies with ≥ 50 rows included",
    x        = "Antibody",
    y        = "% nonconcordant rows") +
  theme_minimal(base_size = 12) +
  expand_limits(y = 100)
ggsave(here("output", "Act3_top15_antibody_noncor_rate.pdf"))

#Since antibodies have much higher rates, verifying top antibody

great_join |> 
  filter(antibody == "HPA002024") |> 
  count(staining_ord, quantity_ord, sort = TRUE)

great_join |> 
  filter(antibody == "HPA002024") |> 
  nrow()
great_join |> 
  filter(antibody == "HPA002024") |> 
  count(Tissue, sort = TRUE)

```
