---
title: "Clean_HPA"
format: html
editor: visual
---

Sample Variability Analysis

## Act 1

Libraries used

```{r}
library(vroom)
library(tidyverse)
library(here)
library(readxl)
library(janitor)
library(irr)
library(patchwork)
```

Data loading

```{r}
files_col_name <- c("gene", "tissue", "antibody","image_link","info_id", "gender", "age", "tissue_detail", "sample_category", "patient_id", "cell_type", "expression_level_category", "reliability", "quantity","location")

all_metadata <- vroom::vroom(
  list.files(here("data","hpa_scraped_metadata_files"),
           full.names = TRUE),
  delim = NULL,
  col_names = files_col_name,
  progress = TRUE,
  id = "source_file")

HUEQ_data <- read.csv(here("data", "10K_Single_Cell_Filtered(in).csv"))

#reliability column in all_metadata is the "intensity" variable from HPA webiste
all_metadata <- all_metadata |> 
  rename(intensity = reliability)


```

Barplot showing exp_level between "tissue"

```{r}
all_metadata|> 
  ggplot(aes(x = tissue, fill = expression_level_category)) +
  geom_bar(position = "fill") +
  labs(title = "Protein Expression Levels by Tissue",
       x = "Tissue Type", y = "Proportion") +
  theme_minimal()
```

### Main figure for counts

Counts for uncomplicated variables

```{r}
all_metadata |> 
  summarise(across(c(gene, tissue, tissue_detail, cell_type, antibody), n_distinct))


```

#### Patient Data

3 Male patients with no age, not very significant choosing to remove outliers since they don't affect downstream analysis

```{r}
patient_histogram <- all_metadata |>
  distinct(patient_id, .keep_all = TRUE) |>
  filter(!is.na(age), !is.na(gender)) |> 
  ggplot() + aes(x = age, fill = gender) + 
  geom_histogram(position = "identity", alpha = 0.6, bins = 30, color = "white") +
  scale_fill_manual(values = c("#86E3CE", "#FA897b")) +  
  theme_minimal() +
  theme(panel.grid.major = element_blank(),   
        panel.grid.minor = element_blank(),  
        axis.line = element_line(color = "grey40"),
        legend.position = "none",            
        plot.title = element_text(face = "bold", size = 14),
        axis.title = element_text(size = 11),
        axis.text = element_text(size = 10)) +
  labs(title = "Age Distribution by Gender",
       x = "Age",
       y = "Count",
       fill = "Gender")


#Gender distribution 
patient_pie <- all_metadata |>
  distinct(patient_id, .keep_all = TRUE) |>
  filter(!is.na(age), !is.na(gender)) |> 
count(gender) |> 
  mutate(percentage = n / sum(n) * 100) |> 
ggplot() + aes(x = "", y = n, fill = gender) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar(theta = "y") + 
  geom_text(aes(label = paste0(round(percentage, 1),"%")),
    position = position_stack(vjust = 0.5),
    size = 4,
    color = "black") +
  scale_fill_manual(values = c("#86E3CE", "#FA897b")) +
  labs(fill = "Gender") +
  theme_minimal() +
  theme(
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9),
    axis.text = element_blank(),
     axis.title = element_blank(),
    panel.grid = element_blank()) 

#Patient output
patient_histogram 
ggsave(here("output","PatientCounts.pdf"), width = 10, height = 7)

patient_pie 
ggsave(here("output","PatientPie.pdf"), width = 5, height = 5)

```

#### Sample Category Distribution

```{r}
figure_sample <- all_metadata |>
  mutate(broad_category = case_when(
    str_detect(sample_category, "^Normal tissue") ~ "Normal Tissue",
    str_detect(sample_category, "^Normal morphology") ~ "Normal Tissue",
    TRUE ~ "Other"))


figure_sample |> 
  count(broad_category) |> 
 mutate(percentage = n / sum(n) * 100,
        broad_category = fct_relevel(broad_category, "Normal Tissue", "Other")) |> 
  ggplot() + aes(x = broad_category, y = percentage, fill = broad_category) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(aes(label = paste0(round(percentage, 1), "%")),
            hjust = -0.1, size = 4, frontface = "bold") +
  scale_y_continuous(limits = c(0, 100), expand = expansion(mult = c(0, 0))) +
  coord_flip() +
  scale_fill_manual(values = c("#86E3CE", "#FA897b")) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x      = element_line(color = "grey40"),
    axis.title       = element_blank(),
    axis.text.y      = element_text(size = 11, face = "bold"),
    axis.text.x      = element_blank(),
    plot.title       = element_text(face = "bold", size = 14)) +
  labs(title = "Sample Category Distribution")
ggsave(here("output","Sample_figure.jpeg"))

```

#### Descriptors Distribution

NA values were dropped due to extremely low quantity (19 counts), #named "not representative" in intensity column

```{r}
Exp_count <- all_metadata |> 
  count(expression_level_category) |> 
  filter(!is.na(expression_level_category)) |> 
   mutate(cat   = factor(expression_level_category, 
                     levels = c("Not detected", "Low", "Medium", "High")),
      pct   = round(n / sum(n) * 100, 1)) |> 
   ggplot() + aes(x = cat, y = pct, fill = cat) +
  geom_col() +
  geom_text(aes(label = paste0(pct, "%\n(n=", n, ")")), hjust = -0.1) + 
  coord_flip(ylim = c(0, 60)) +
  scale_fill_manual(values = c(
  "Not detected" = "#86e3ce",
  "Low"  = "#d0e6a5",
  "Medium" = "#ffdd94",
  "High" = "#fa897b")) +
  scale_y_continuous(breaks = seq(0, 50, by = 10)) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),   
        panel.grid.minor = element_blank(),  
        axis.line = element_line(color = "grey40"),
        legend.position = "none",            
        plot.title = element_text(face = "bold", size = 14),
        axis.title = element_text(size = 11),
        axis.text = element_text(size = 10)) +
  labs(title = "Expression level Distribution",
       y = "(%) of Dataset", 
       x = "Expression Level") 
ggsave(here("output","Expression_Dist.pdf"))

intensity_counts <- all_metadata |> 
count(intensity) |> 
  filter(!intensity == "Not representative") |> 
   mutate(cat   = factor(intensity, 
                     levels = c("Negative", "Weak", "Moderate", "Strong")),
      pct   = round(n / sum(n) * 100, 1)) |> 
   ggplot() + aes(x = cat, y = pct, fill = cat) +
  geom_col() +
  geom_text(aes(label = paste0(pct, "%\n(n=", n, ")")), hjust = -0.1) + 
  coord_flip(ylim = c(0, 60)) +
  scale_fill_manual(values = c(
  "Negative" = "#86e3ce",
  "Weak"= "#d0e6a5",
  "Moderate"= "#ffdd94",
  "Strong" = "#fa897b")) +
  scale_y_continuous(breaks = seq(0, 50, by = 10)) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),   
        panel.grid.minor = element_blank(),  
        axis.line = element_line(color = "grey40"),
        legend.position = "none",            
        plot.title = element_text(face = "bold", size = 14),
        axis.title = element_text(size = 11),
        axis.text = element_text(size = 10)) +
  labs(title = "Staining Intensity Distribution",
       y = "(%) of Dataset", 
       x = "Intensity Category") 

Exp_count + intensity_counts
ggsave(here("output","Descriptor_count_figures.pdf"))

```

## Act 2

Cross tables, using tabyl to make how many times x variable appear in each tissue. Used for descriptive structre

```{r}
#Gene x tissue frequency and proportions 
cross_gene_tissue <- all_metadata |> 
  tabyl(gene, tissue) |> 
  adorn_totals("both") 

cross_gene_tissue_prop <- all_metadata |> 
  tabyl(gene, tissue) |> 
  adorn_percentages("row") |> 
  adorn_pct_formatting()

#Antibody x Intensity frequency and proportions 
cross_anti_intensity <- all_metadata |> 
  tabyl(antibody, intensity) |> 
  adorn_totals("both")
cross_anti_intensity_prop <- all_metadata |> 
  tabyl(antibody, intensity) |> 
  adorn_percentages("row") |> 
  adorn_pct_formatting()


#an example of three way cros: intensity x patient x sample category 
cross3way_rel_patient_sampleCat <- all_metadata |> 
  group_by(intensity) |> 
  summarise(
    table = list(tabyl(cur_data(), patient_id, sample_category)))
```

"Not detected" category profiling per patient and gene: Classifies each patient by whether their entries are entirely "Not detected", mixed or clean.

```{r}
#Purely per patient
all_metadata |>
  group_by(patient_id) |>
  summarise(n_total = n(),
    has_not_detected = any(expression_level_category == "Not detected", na.rm = TRUE),
    has_other = any(expression_level_category != "Not detected", na.rm = TRUE),
    .groups = "drop") |>
  mutate(category = case_when(
      has_not_detected & has_other ~ "mixed (not detected + other result)",
      has_not_detected & !has_other ~ "only not detected",
      !has_not_detected & has_other ~ "no not detected (only real results)",
      TRUE ~ "other"))

#Per Patient x gene 
all_metadata |>
  group_by(patient_id, gene) |>
  summarise(n_tests = n(),
    has_not_detected = any(expression_level_category == "Not detected",
                           na.rm = TRUE),
    has_other = any(expression_level_category != "Not detected",
                    na.rm = TRUE),
    n_not_detected = sum(expression_level_category == "Not detected", 
                         na.rm = TRUE),
    n_other = sum(expression_level_category != "Not detected", 
                  na.rm = TRUE),
    .groups = "drop") |>
  mutate(category = case_when(
      has_not_detected & !has_other ~ "only not detected (no retest)",
      has_not_detected & has_other ~ "not detected and a retest",
      !has_not_detected ~ "never not detected"))
```

### Data Merge & Exploration

```{r}
mergeable_metadata <- all_metadata |>
  mutate(tissue   = str_to_title(tissue),
         tissue   = str_remove(tissue, "\\+.*"),
         antibody = str_remove(antibody, "Antibody ")) |>
  rename(Protein        = gene,
         Patient_ID     = patient_id,
         Tissue         = tissue,
         Antibody       = antibody,
         HPA_Staining   = expression_level_category,
         HPA_Intensity  = intensity)                  

great_join <- HUEQ_data |>
  left_join(mergeable_metadata |>
      select(Antibody, Protein, Patient_ID, Tissue, HPA_Staining, HPA_Intensity),
    by = c("Antibody", "Protein", "Patient_ID", "Tissue")) |> 
  distinct(Antibody, Protein, Patient_ID, Tissue, Quantity, .keep_all = TRUE) |> 
  mutate(quantity_ord = factor(Quantity, 
                               levels = c("None", "<25%", "75%-25%", ">75%"), 
                               ordered = TRUE), 
         hpa_staining_ord = factor(HPA_Staining, 
                               levels = c("Not detected", "Low", "Medium", "High"),
                               ordered = TRUE), 
         hpa_intensity_ord = factor(HPA_Intensity, 
                                levels = 
                                  c("Negative", "Weak", "Moderate", "Strong"),
                                ordered = TRUE))
# 3 duplicate matches removed (with distinct): HPA stores sub-regional tissue entries 
# (e.g. upper vs lower stomach) that collapse to the same Tissue label causing one HUEQ row to match multiple all_metadata entries
```

## ACT 3

### Weighted Kappa: Interrater agreement

For the purposes of this analysis the "raters" are the ordinal descriptors or labels: Staining (Expression Level Category), Intensity (Staining Intensity) & Quantity.

```{r}
kappa_staining<- great_join |> 
  filter(!is.na(HPA_Staining)) |> 
  transmute(rater1 = as.numeric(hpa_staining_ord),
            rater2 = as.numeric(quantity_ord)) |> 
  kappa2(weight = "squared")
kappa_staining

kappa_intensity <- great_join |> 
  filter(!is.na(HPA_Intensity)) |> 
  transmute(rater1 = as.numeric(hpa_intensity_ord),
            rater2 = as.numeric(quantity_ord)) |> 
  kappa2(weight = "squared")
kappa_intensity

```

### Heat maps

HPA labels (Staining & Intensity) vs HPA Quantity (derived from HUEQ)

```{r}
#Agreement table 
concordance_StainQuant <- great_join |> 
  tabyl(hpa_staining_ord, quantity_ord) |> 
  adorn_totals("both") |> 
  adorn_percentages("row") |> 
  adorn_pct_formatting()
#First Heat map 
Heatmap_A <- great_join |> 
  filter(!is.na(HPA_Staining)) |> 
  count(hpa_staining_ord, quantity_ord) |> 
  group_by(hpa_staining_ord) |> 
  mutate(prop = n / sum(n)) |> 
  ungroup() |> 
  ggplot() + aes(x = hpa_staining_ord, y = quantity_ord, fill = prop) +
  geom_tile() + geom_text(aes(label = scales::percent(prop, accuracy = 1)),
                          colour = "white", frontface = "bold") +
  labs(title = "HPA Expression Label vs HPA Quantity Label",
       subtitle = "Inter-label agreement across 16,643 observations",
       y = "HPA Quantity Label, derived from HUEQ (%)", 
       x = "HPA Expression Level Category", 
       fill = "Row proportion") 
#Second Heat map
Heatmap_B <- great_join |> 
  filter(!is.na(HPA_Intensity)) |> 
  count(hpa_intensity_ord, quantity_ord) |> 
  group_by(hpa_intensity_ord) |> 
  mutate(prop = n / sum(n)) |> 
  ungroup() |> 
  ggplot() + aes(x = hpa_intensity_ord, y = quantity_ord, fill = prop) +
  geom_tile() + geom_text(aes(label = scales::percent(prop, accuracy = 1)),
                          colour = "white", frontface = "bold") +
  labs(title = "HPA Staining Intensity  vs HPA Quantity Label",
       subtitle = "Inter-label agreement across 16,643 observations",
       y = "HPA Quantity Label, derived from HUEQ (%)", 
       x = "HPA Staining Intensity", 
       fill = "Row proportion") 
#Appendix Figure
Heatmap_C <- great_join |> 
  filter(!is.na(HPA_Intensity)) |> 
  count(hpa_intensity_ord, hpa_staining_ord) |> 
  group_by(hpa_intensity_ord) |> 
  mutate(prop = n / sum(n)) |> 
  ungroup() |> 
  ggplot() + aes(x = hpa_intensity_ord, y = hpa_staining_ord, fill = prop) +
  geom_tile() + geom_text(aes(label = scales::percent(prop, accuracy = 1)),
                          colour = "white", frontface = "bold") +
  labs(title = "HPA Staining Intensity  vs HPA Quantity Label",
       subtitle = "Inter-label agreement across 16,643 observations",
       y = "HPA Quantity Label, derived from HUEQ (%)", 
       x = "HPA Staining Intensity", 
       fill = "Row proportion") 


```

### Per Row agreement Agreement

Between hpa_staining and quantity, as compared in Kappa test. HPA Intensity's "Strong" label should have higher agreement between HPA_Staining and Quantity than "Weak" label entries.

```{r}
agreement_test <- great_join |>
  mutate(concordant = case_when(is.na(HPA_Staining) ~ NA,
         quantity_ord == "None" & hpa_staining_ord == "Not detected" ~ TRUE,
         quantity_ord == "<25%" & hpa_staining_ord == "Low" ~ TRUE,
         quantity_ord == "75%-25%" & hpa_staining_ord == "Medium" ~ TRUE,
         quantity_ord == ">75%" & hpa_staining_ord == "High" ~ TRUE,
         TRUE  ~ FALSE )) |> 
  filter(!is.na(HPA_Intensity), !is.na(concordant)) |> 
  group_by(HPA_Intensity) |>
  summarise( n = n(),
             n_concordant = sum(concordant),
             pct_concordant = round(n_concordant / n * 100, 1),
             .groups = "drop") |> 
  arrange(desc(pct_concordant))

#Visualization of Agreement 

agreement_figure <- agreement_test |> 
  ggplot() + aes(x = reorder(HPA_Intensity, pct_concordant), 
                 y = pct_concordant, 
                 fill = pct_concordant) + geom_col() +
  geom_text(aes(label = paste0(pct_concordant,
                             "%\n(n=", n, ")")), 
            hjust = -0.1) + 
  coord_flip(ylim = c(0, 110)) + 
  labs( title = "Agreement Rate Between HPA Expression Level and Quantity by Intensity",
        subtitle = "Does HPA Intensity predict consistency between the other two descriptors?",
        x = "HPA Staining Intensity",
        y = "% rows where HPA Expression label agrees with Quantity") 

```

### Disagreement

In which cases do the labels not agree

```{r}
#Per Tissue 
disagreement_figure_tissue <- great_join |> 
   mutate(concordant = case_when(is.na(HPA_Staining) ~ NA,
         quantity_ord == "None" & hpa_staining_ord == "Not detected" ~ TRUE,
         quantity_ord == "<25%" & hpa_staining_ord == "Low" ~ TRUE,
         quantity_ord == "75%-25%" & hpa_staining_ord == "Medium" ~ TRUE,
         quantity_ord == ">75%" & hpa_staining_ord == "High" ~ TRUE,
         TRUE  ~ FALSE )) |> 
  filter(!is.na(HPA_Intensity), !is.na(concordant)) |>
  group_by(Tissue) |>
  summarise(n_total        = n(),
            n_nonconcordant = sum(!concordant, na.rm = TRUE),
            pct_nonconcordant = round(n_nonconcordant / n_total * 100, 1),
            .groups = "drop") |>
  filter(n_total >= 50) |>       
  slice_max(pct_nonconcordant, n = 15) |>
  ggplot() + aes(x = reorder(Tissue, pct_nonconcordant), 
                 y = pct_nonconcordant,
                 fill = pct_nonconcordant) +
  geom_col() +
  coord_flip() +
  scale_fill_distiller(palette = "Spectral", 
                       direction = -1,      
                       name = "% disagreement") +
  geom_text(aes(label = paste0(pct_nonconcordant, "%")),
            hjust = -0.2, size = 3.5) +
  labs(
    title    = "Top 15 Tissues by Inter-descriptor Disagreement Rate",
    subtitle = "Only tissues with ≥ 50 rows included",
    x        = "Tissue",
    y        = "% disagreement rows") +
  theme_minimal(base_size = 12) +
  expand_limits(y = 100)

#Per Antibody rate 
disagreement_figure_antibody <- great_join |> 
   mutate(concordant = case_when(is.na(HPA_Staining) ~ NA,
         quantity_ord == "None" & hpa_staining_ord == "Not detected" ~ TRUE,
         quantity_ord == "<25%" & hpa_staining_ord == "Low" ~ TRUE,
         quantity_ord == "75%-25%" & hpa_staining_ord == "Medium" ~ TRUE,
         quantity_ord == ">75%" & hpa_staining_ord == "High" ~ TRUE,
         TRUE  ~ FALSE )) |> 
  filter(!is.na(HPA_Intensity), !is.na(concordant)) |>
  group_by(Antibody) |>
  summarise(n_total        = n(),
            n_nonconcordant = sum(!concordant, na.rm = TRUE),
            pct_nonconcordant = round(n_nonconcordant / n_total * 100, 1),
            .groups = "drop") |>
  filter(n_total >= 50) |>       
  slice_max(pct_nonconcordant, n = 15) |>
  ggplot() + aes(x = reorder(Antibody, pct_nonconcordant), 
                 y = pct_nonconcordant,
                 fill = pct_nonconcordant) +
  geom_col() +
  coord_flip() +
  scale_fill_distiller(palette = "Spectral", 
                       direction = -1,      
                       name = "% disagreement") +
  geom_text(aes(label = paste0(pct_nonconcordant, "%")),
            hjust = -0.2, size = 3.5) +
  labs(
    title    = "Top 15 Antibodies by Inter-descriptor Disagreement Rate",
    subtitle = "Only Antibodies with ≥ 50 rows included",
    x        = "Antibody",
    y        = "% disagreement rows") +
  theme_minimal(base_size = 12) +
  expand_limits(y = 100)

```
